---
title: "Storm Weather Analysis"
author: "Richie Liew"
date: "Nov 15, 2016"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=13, fig.height=9, fig.path='Figures/', echo=TRUE, warning=FALSE, message=FALSE)
```

### Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

### Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* Storm Data [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service Storm Data Documentation

* National Climatic Data Center Storm Events FAQ

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

### Data Analysis

Library Required:
```{r}
library(ggplot2)
```

#### 1. Read Storm Database.  Compress file are downloaded to ./data folder

```{r}
storm_db <- read.csv(file="./data/StormData.csv.bz2", header=T)
```

#### 2. Load Storm Database with the require varaibles.

Here are the require varabiles.

LABEL | DESCRIPTION
---|-----------------------------------------------------------------------------------
DATE_BEGIN | Event Begin Date
EVENT_TYPE | Event Type
FATALITIES | Number of reported fatalities caused by the event.
INJURIES | Number of reported injuries caused by the event.
PROPDMG | Amount of property damage caused by the event.
CROPDMG | Amount of crop damage caused by the event.
PROPDMGEXP | The dollar (USD) amount of property damage caused by the event.
CROPDMGEXP | The dollar (USD) amount of crop damage caused by the event.

```{r}
col_set <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
storm_set <- storm_db[col_set]
str(storm_set)
head(storm_set)
```

#### 3. Cleaning the Data
Property Damage Exponent Unique Value
```{r}
unique(storm_set$PROPDMGEXP)
```

Crop Damage Exponent Unique Value
```{r}
unique(storm_set$CROPDMGEXP)
```

Property Damage Data:
```{r}  
storm_set$PROPDMGEXP <- as.character(storm_set$PROPDMGEXP)
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "K"] <- 1000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "M"] <- 1000000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == ""] <- 1
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "B"] <- 1000000000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "m"] <- 1000000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "0"] <- 1
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "5"] <- 100000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "6"] <- 1000000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "4"] <- 10000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "2"] <- 100
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "3"] <- 1000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "h"] <- 100
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "7"] <- 10000000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "H"] <- 100
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "1"] <- 10
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "8"] <- 100000000
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "+"] <- 0
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "-"] <- 0
storm_set$PROPDMGEXP[storm_set$PROPDMGEXP == "?"] <- 0
storm_set$PROPDMGEXP <- as.integer(storm_set$PROPDMGEXP)
storm_set$PROPDMGVALUE <- storm_set$PROPDMG * storm_set$PROPDMGEXP
```

```{r}
storm_set$CROPDMGEXP <- as.character(storm_set$CROPDMGEXP)
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "M"] <- 1000000
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "K"] <- 1000
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "m"] <- 1000000
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "B"] <- 1000000000
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "0"] <- 1
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "k"] <- 1000
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "2"] <- 100
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == ""] <- 1
storm_set$CROPDMGEXP[storm_set$CROPDMGEXP == "?"] <- 0
storm_set$CROPDMGEXP <- as.integer(storm_set$CROPDMGEXP)
storm_set$CROPDMGVALUE <- storm_set$CROPDMG * storm_set$CROPDMGEXP
```

### Results

#### Top 10 Death:
```{r}
fatality <- aggregate(FATALITIES ~ EVTYPE, storm_set, FUN = sum)
```

Here are the top 10 death.

```{r}
TOP10DEAD <- fatality[order(-fatality$FATALITIES), ][1:10,]
TOP10DEAD
```

#### Top 10 Injuries:

```{r}
injury   <- aggregate(INJURIES ~ EVTYPE, storm_set, FUN = sum)
```

Here are the top 10 injuries.

```{r}
TOP10INJ <-injury[order(-injury$INJURIES), ][1:10,]
TOP10INJ
```

```{r}
TOP5DEAD <- fatality[order(-fatality$FATALITIES), ][1:5,]
TOP5INJ <-injury[order(-injury$INJURIES), ][1:5,]
```

```{r}
par(mfrow=c(2,2))
barplot(TOP10DEAD$FATALITIES,names.arg=TOP10DEAD$EVTYPE,space=0.2,horiz=TRUE,col="cyan",
        xlab="Fatalities",xlim=c(0,6000),cex.axis=0.7,ylab="Event Type",ylim=c(0,10),cex.names=0.5,
        main="Top 10 Causes of Total Fatalities")
barplot(TOP10INJ$INJURIES,names.arg=TOP10INJ$EVTYPE,space=0.2,horiz=TRUE,col="green",
        xlab="Injuries",xlim=c(0,90000),cex.axis=0.7,ylab="Event Type",ylim=c(0,10),cex.names=0.5,
        main="Top 10 Total Causes of Injuries") 
barplot(TOP5DEAD$FATALITIES, names.arg=TOP5DEAD$EVTYPE,space=0.2,horiz=TRUE,col="red",
        xlab="Fatalities",xlim=c(0,6000),cex.axis=0.7,ylab="Event Type",ylim=c(0,5),cex.names=0.5,
        main="Top 5 Causes of Total Fatalities")
barplot(TOP5INJ$INJURIES, names.arg = TOP5INJ$EVTYPE,space=0.2,horiz=TRUE,col="blue",
        xlab="Injuries",xlim=c(0,90000),cex.axis=0.7,ylab="Event Type",ylim=c(0,5),cex.names=0.5,
        main="Top 5 Total Causes of Injuries")
```        
        
```{r}
propertydmg <- aggregate(PROPDMGVALUE ~ EVTYPE, storm_set, FUN = sum)
cropdmg   <- aggregate(CROPDMGVALUE ~ EVTYPE, storm_set, FUN = sum)
head(propertydmg)
head(cropdmg)
```

### Economic Consequences

To analyze the top contributing weather events by total cost of damage, both property and crop damage amounts are totaled for ranking.




